---
title: "COVID-19 Impact on Seattle Property Values"
author: "Justin Gao"
format: 
  revealjs:
    embed-resources: true
    theme: default
    slide-number: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

house_prices <- read_csv("seattle_house_prices.csv", show_col_types = FALSE)
building_permits <- read_csv("seattle_building_permits.csv", show_col_types = FALSE)
```

## Introduction

**Justin Gao**

INFO 201

Section AD

Unfortunately it is just me working on this project by myself. I believe my partner dropped the class

# Research Question

---

## Big Picture Question

**How did the COVID-19 pandemic affect residential property values and development activity in Seattle?**

<br>

**Sub-questions:**

1. Did house prices increase or decrease during COVID?
2. How did price growth rates change?
3. Did construction activity slow down?
4. Were single-family and multifamily housing affected differently?

# Background

---

## Previous Research

**Source:** Francke, M. & Korevaar, M. (2021). "Housing markets in a pandemic: Evidence from historical outbreaks." *Journal of Urban Economics*

<br>

**Important takeaways:**

- Historical pandemics showed temporary price declines
- Low interest rates during COVID created unique conditions
- Remote work increased demand for housing space



**Why this matters:** Helps us understand if Seattle followed historical patterns

# Data

---

## Dataset 1: House Prices

**Source:** Federal Housing Finance Agency via FRED

**Link:** https://fred.stlouisfed.org/series/ATNHPIUS42644Q




**Sample:** 31 quarterly observations, Q1 2018 - Q3 2025

**What it measures:** House price index for Seattle metro area

**Population:** Residential properties with conforming mortgages

**Who created it:** U.S. Federal Housing Finance Agency

---

## Dataset 2: Building Permits

**Source:** City of Seattle Open Data Portal

**Link:** https://data.seattle.gov/Built-Environment/Building-Permits/76t5-zqzr




**Sample:** 34 permits from 2018-2024

**What it measures:** Construction permits with its type, cost, housing units

**Population:** All the construction projects requiring permits in Seattle

**Who created it:** Seattle Department of Construction & Inspections

# Data Wrangling

---

## Steps Taken

**1. Date Processing (Novel Tool)**

Used ymd(), year(), quarter() to work with dates



**2. COVID Period Classification (Basic: mutate)**

Created Pre-COVID, During COVID, Post-COVID periods



**3. Summary Statistics (Basic: group_by + summarise)**

Calculated averages by period



**4. Joining Datasets (Intermediate)**

Combined house prices and permits by time period

---

## Date Tool Example

```{r}
#| echo: true
#| eval: false

house_prices <- house_prices %>%
  mutate(
    DATE = ymd(DATE),
    Year = year(DATE),
    Quarter = quarter(DATE),
    COVID_Period = case_when(
      Year < 2020 ~ "Pre-COVID",
      Year >= 2020 & Year <= 2021 ~ "During COVID",
      TRUE ~ "Post-COVID"
    )
  )
```

**Why needed:** Can't do time analysis without proper dates

**Resource:** https://r4ds.had.co.nz/dates-and-times.html

# Analysis

---

## Question 1: Price Trends

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5


house_prices_clean <- house_prices %>%
  mutate(
    DATE = ymd(DATE),
    Year = year(DATE),
    COVID_Period = case_when(
      Year < 2020 ~ "Pre-COVID",
      Year >= 2020 & Year <= 2021 ~ "During COVID",
      TRUE ~ "Post-COVID"
    ),
    COVID_Period = factor(COVID_Period, levels = c("Pre-COVID", "During COVID", "Post-COVID"))
  )



ggplot(house_prices_clean, aes(x = DATE, y = ATNHPIUS42644Q, color = COVID_Period)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  labs(title = "Seattle House Price Index Over Time",
       x = "Date",
       y = "Price Index",
       color = "Period") +
  theme_minimal() +
  scale_color_manual(values = c("Pre-COVID" = "blue", 
                                  "During COVID" = "orange", 
                                  "Post-COVID" = "green"))
```

**Finding:** 

- From 2020 to 2021, price jumped from 452 to 549, but from 2018 to 2019, price had a steady growth from 408 to 448

- Prices jumped 21% during COVID vs 10% growth before

---

## Question 2: Growth Rates

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

house_prices_growth <- house_prices_clean %>%
  arrange(DATE) %>%
  mutate(Growth_Rate = (ATNHPIUS42644Q - lag(ATNHPIUS42644Q)) / lag(ATNHPIUS42644Q) * 100) %>%
  filter(!is.na(Growth_Rate))

ggplot(house_prices_growth, aes(x = COVID_Period, y = Growth_Rate, fill = COVID_Period)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter() +
  labs(title = "Quarterly Growth Rates by Period",
       x = "Period",
       y = "Growth Rate (%)") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Pre-COVID" = "blue", 
                                "During COVID" = "orange", 
                                "Post-COVID" = "green")) 
```

**Finding:** During COVID had highest growth (3.5% per quarter)

---

## Question 3: Development Activity

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5


building_permits_final <- building_permits %>%
  mutate(
    AppliedDate = ymd(AppliedDate),
    AppliedYear = year(AppliedDate),
    COVID_Period = case_when(
      AppliedYear < 2020 ~ "Pre-COVID",
      AppliedYear >= 2020 & AppliedYear <= 2021 ~ "During COVID",
      TRUE ~ "Post-COVID"
    ),
    COVID_Period = factor(COVID_Period, levels = c("Pre-COVID", "During COVID", "Post-COVID"))
  )




permit_summary <- building_permits_final %>%
  group_by(COVID_Period, PermitClass) %>%
  summarise(Count = n(), .groups = 'drop')





ggplot(permit_summary, aes(x = COVID_Period, y = Count, fill = PermitClass)) +
  geom_col(position = "dodge") +
  labs(title = "Building Permits by Type and Period",
       x = "Period",
       y = "Number of Permits",
       fill = "Type") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 4)
```

**Finding:** Development mostly stayed stable during COVID and then increased after

---

## Question 4: Housing Types

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

cost_summary <- building_permits_final %>%
  filter(PermitClass != "Commercial") %>%
  group_by(COVID_Period, PermitClass) %>%
  summarise(Avg_Cost = mean(EstProjectCost, na.rm = TRUE), .groups = 'drop')

ggplot(cost_summary, aes(x = COVID_Period, y = Avg_Cost/1000000, fill = PermitClass)) +
  geom_col(position = "dodge") +
  labs(title = "Average Project Costs by Housing Type",
       x = "Period",
       y = "Cost (Millions $)",
       fill = "Type") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Single Family / Duplex" = "steelblue", 
                                "Multifamily" = "coral")) +
  geom_text(aes(label = paste0("$", round(Avg_Cost/1000000, 1), "M")), 
            position = position_dodge(width = 0.9), vjust = -0.5, size = 4)
```

**Finding:** Both increased, but multifamily dropped during COVID then surged

# Advanced Analysis

---

## Joining Datasets (Intermediate Tool)

```{r}
#| echo: true

price_by_period <- house_prices_clean %>%
  group_by(COVID_Period) %>%
  summarise(Avg_Price_Index = mean(ATNHPIUS42644Q), .groups = 'drop')

permits_by_period <- building_permits_final %>%
  group_by(COVID_Period) %>%
  summarise(Total_Permits = n(), .groups = 'drop')

combined <- price_by_period %>%
  inner_join(permits_by_period, by = "COVID_Period")
```

**Why needed:** I need to do this to see relationships between prices and building

**Alternative:** You could analyze separately, but joining still shows connections

**Resource:** https://r4ds.had.co.nz/relational-data.html

---

## Price vs Development Relationship

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

ggplot(combined, aes(x = Avg_Price_Index, y = Total_Permits, color = COVID_Period)) +
  geom_point(size = 7) +
  geom_text(aes(label = COVID_Period), vjust = -1.2, size = 4) +
  labs(title = "House Prices vs Building Permits",
       x = "Average Price Index",
       y = "Total Permits") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Pre-COVID" = "blue", 
                                  "During COVID" = "orange", 
                                  "Post-COVID" = "green")) +
  theme(legend.position = "none")
```

**Finding:** High prices during COVID didn't immediately increase building

# Conclusions

---

## Main Findings

**COVID drastically increased housing prices:**




1. Prices jumped about 21% during pandemic (2020-2021)

2. Quarterly growth rates were the highest during COVID

3. Construction stayed steady, and continued to increase post-COVID

4. Single-family costs rose steadily, however the multifamily costs were volatile

5. High prices didn't immediately lead to more building



---

## Limitations

**What we couldn't address:**


1. Small sample size (cuz only 34 permits)

2. Permits are Seattle-only and prices are metro-wide

3. No data on interest rates or employment

4. Can't prove causation only patterns

5. Missing neighborhood-level details



**Future work:** Add economic data, compare other cities, study neighborhoods

# Summary

---

## Tools & Data Used

**Data Sources:**

- House prices: https://fred.stlouisfed.org/series/ATNHPIUS42644Q
- Building permits: https://data.seattle.gov/Built-Environment/Building-Permits/76t5-zqzr

<br>

**Basic Tools:** mutate, filter, group_by, summarise, arrange

**Intermediate Tool:** inner_join

**Novel Tool:** Date functions (ymd, year, quarter)




**Visualizations:** Line graph, box plot, grouped bar chart

---

# Thank You

**By: Justin Gao** 
